---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# *PointedSDMs* workshop material

This page is dedicated for the material used in the *PointedSDMs* workshop. It contains all the presentations, vignettes and data required to run the analyses.

To install the *R* package, please use the following script:

```{r Install, eval=FALSE}
install.packages('PointedSDMs')
#or
devtools::install_github('PhilipMostert/PointedSDMs@main')

```

You will also need INLA, which you can install from [here](https://www.r-inla.org/download-install), which recommends using this code:


```{r InstallINLA, eval=FALSE}
install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
```

# Approximate timings

10:00 - 10:20: Introduction & first presentation

10:20 - 11.05: First exercise (most time taken installing pkg & INLA)

11:05 - 11:40: Second presentaiton, but had a LOT of questions

11:40-12:10: Second exercise

11:10-12:20: Third presentation

12:20-12:25: Break
