---
title: "Spatio-temporal"
author: "Philip S. Mostert"
date: "2024-06-17"
output: pdf_document
bibliography: References.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width=8, 
                      fig.height=5)
```

This vignette is dedicated to illustrating a spatio-temporal ISDM analysis using the *PointedSDMs R* package. The analysis presented here is loosely adapted from a similar analysis presented in [@seaton2024].

We aim to estimate maps of log-intensity across time for *Pyronia tithonus* (common name: *Gatekeeper*) across England, Scotland and Wales. Two datasets are considered: the *UK butterfly monitoring scheme* (*ukbms*), which are collected at regularly surveyed transects across the UK. The data comes in the form of an abundance dataset, however for this analysis we treated them as detection/non-detection data. The other dataset considered came from the *British Trust for Ornithology plus partner organisations* (*bto*). These data are collected by citizen scientists, and we therefore treated them as presence only data.

We load the packages required by the vignette:

```{r Load packages}
library(sf)
library(dplyr) 
library(tidyr)
library(ggplot2)
library(INLA)
library(PointedSDMs)
```

And define a coordinate reference system to be used:

```{r Proj}
km_projection <- "+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +ellps=airy +units=km +no_defs"
```

The data used in this analysis contains information from the year 2000 all the way to the year 2018. Estimating a model with too many time periods may take a long time to estimate. Therefore the time periods considered may be filtered using the following script. Note that estimating a model with three time periods as chosen here may still take a long while to estimate.

```{r Choose time}
yearBegin <- 2006 #2000
yearEnd <- 2008 #2018
```

We next load in the data used in the analysis, and a *sf* object for the study region, which has been simplified slightly in order to produce a better mesh:

```{r Load ukbms}
ukbms <- readRDS(file = 'Data/Vignette_temporal/ukbms.rds')
```

```{r Load bto}
bto <- readRDS(file = 'Data/Vignette_temporal/bto.rds')
```

```{r Load GB}
gb <- readRDS(file = 'Data/Vignette_temporal/gb.rds')
```

We then subset the data based on the chosen years.

```{r Subset time}
ukbms <- ukbms %>%
  filter(Year >= yearBegin & Year <= yearEnd)

bto <- bto %>%
  filter(Year >= yearBegin & Year <= yearEnd)
```

An *inla.mesh* object is created to estimate the spatial effect. We use a coarse mesh to speed up the run time of the models.

```{r Mesh}
mesh <- inla.mesh.2d(boundary = INLA::inla.sp2segment(gb), 
                     max.edge = c(5,30) * 4,
                     cutoff = 2,
                     offset = 2,
                     crs = km_projection)
plot(mesh)
```

We set up the model using the function `startISDM`, and specify the argument *temporalName* to set up a temporal analysis. The argument is the name of the column which contains information on the temporal variable, and is required to be standardized across datasets. We create a plot of the data to see how the data is spread across the map.

```{r startISDM}
model_setup <- startISDM(ukbms, bto, 
                         Projection = km_projection, Mesh = mesh,
                         Boundary = gb, responsePA = 'Presence', 
                         temporalName = 'Year')
plot(model_setup)
```

A second spatial effect is added to the *bto* dataset to account for the spatial biases inherent in these data.

```{r addBias}
model_setup$addBias(datasetNames = 'bto')

```

Priors for the spatial effects and the hyperparameters of the model are then set.

```{r Priors}
model_setup$specifySpatial(sharedSpatial = TRUE,
                           prior.range = c(100, 0.01),
                           prior.sigma = c(5, 0.01))

model_setup$specifySpatial(Bias = TRUE,
                           prior.range = c(100, 0.01),
                           prior.sigma = c(5, 0.01))

ar1_hyper <- list(model = 'ar1', hyper = list(theta1 = list(prior = "pc.prec", 
                                                            param = c(0.25, 0.01)),
                                 rho = list(prior = "pc.cor1", 
                                            param = c(0.8, 0.8))))

model_setup$specifyRandom(temporalModel = ar1_hyper)
```

Estimation of the model may be completed `fitISDM`.

```{r fitISDM}
TempModel <- fitISDM(model_setup, options = list(num.threads =2,
                                  verbose = TRUE, 
                                  control.inla = list(
                                  int.strategy = 'eb',
                                  strategy = "adaptive",
                                  diagonal = 1)))
summary(TempModel)
```

And finally we create temporal predictions of the model using `predict`,

```{r Predict}
pred <- predict(TempModel, mesh = mesh, 
                mask = gb, predictor = TRUE)
plot(pred)
```

And temporal predictions of the bias component in a similar fashion.

```{r Predict Bias}
predBias <- predict(TempModel, mesh = mesh,
                    mask = gb, bias = TRUE)
plot(predBias)
```
