---
title: 'Presentation 2:'
subtitle: "Introduction to *PointedSDMs*\n\n'ISEC 2024 -- Swansea'"
author: 'Philip Mostert'
institute: 'Norwegian University of Science and Technology'
date: '13 July 2024'
output: beamer_presentation
bibliography:  "`r here::here('References.bib')`"
biblio-style: authoryear
header-includes:
  - \let\verbatim\undefined
  - \let\verbatimend\undefined
  - \usepackage{listings}
  - \lstnewenvironment{verbatim}{\lstset{breaklines=true,basicstyle=\ttfamily\footnotesize}}{}
---

```{r Load PointedSDMs, echo = FALSE, warning = FALSE, message = FALSE}

library(knitr)
library(PointedSDMs)
opts_chunk$set(tidy.opts=list(width.cutoff=20),tidy=TRUE)

```

## Introduction

-   Previous presentation introduced *ISDMs.*

-   In this presentation we will present an *R* package to estimate these models.

## Introduction to integrated modelling

-   Species distribution modelling has many applications in ecology and is now a mature discipline.
-   Previously, the approach used was to favour one dataset, and either discard the other datasets or use them in secondary analyses.
-   However integrated species distribution models (ISDMs) have emerged -- a method to model multiple datasets in a common statistical framework.

## Process model

\small

The underlying process model is a log-Gaussian cox process (LGCP) with a spatially varying intensity function, $\lambda(s) = exp\{\eta(s)\}$, where: $$ \eta(s) = \alpha_0 + \sum_{k = 1}^p\beta_k X_k(s) + \zeta(s),$$ where:

-   $\alpha_0$ is an intercept term,

-   $\beta_k$ is the $k^{th}$ coefficient for the $k^{th}$ environmental covariate, which varies in space (denoted by the index, $s$),

-   $\zeta(s)$ is a zero-mean Gaussian random field determined by a Matèrn covariance function.

## Observation models

\small

-   Observation models are chosen based on the underlying sampling protocols of a given dataset.

| Data type      | Distribution                     |
|:---------------|:---------------------------------|
| Present absent | Binomial (with a *cloglog link*) |
| Present only   | Thinned Poisson                  |
| Count data     | Poisson                          |
| Marks/traits   | Variety                          |

## *PointedSDMs* *R* package

-   Package is designed to simplify the construction of ISDMs using a joint-likelihood framework.
-   Wrapper around *R-INLA* and *inlabru* packages.
-   Manuscript available in *Methods in Ecology and Evolution:* @mostert2023pointedsdms

## Installation of the package

-   Install from either *CRAN* or *Github.*

```{r install package, eval=FALSE}
install.packages('PointedSDMs')
#or
devtools::install_github('PhilipMostert/PointedSDMs')
```

## Package structure

-   Initalize model,

-   Specify model,

-   Estimate model,

-   Predict and plot,

-   Cross-validation.

# Initializing model

## `startISDM`

\small

```{r startISDM}
args(startISDM)
```

## `startSpecies`

\small

```{r startSpecies}
args(startSpecies)
```

## `startMarks`

\small

```{r startMarks}
args(startMarks)
```

# Specifying model

## Available slot functions

| Name                  | Use                                            |
|-----------------------|------------------------------------------------|
| `$plot()`             | Plot data                                      |
| `$addBias()`          | Add second spatial effect                      |
| `$updateFormula()`    | Update a likelihood's formula                  |
| `$changeComponents()` | Update model components                        |
| `$priorsFixed()`      | Specify priors for the fixed effects           |
| `$specifySpatial()`   | Specify the spatial effects                    |
| `$changeLink()`       | Change the link function of a process          |
| `$spatialBlock()`     | Initiate spatial-block cross-validation        |
| `$addSamplers()`      | Change the integration domain for a likelihood |
| `$specifyRandom()`    | Specify the priors for other random effects    |

## Information about slot functions

-   Obtain documentation through the `.`\$`help()` function.

\small

```{r slot functions, eval = FALSE}
obj <- startISDM(...)
obj$help()
```

# Estimate the model

## `fitISDM`

```{r fitISDM}
args(fitISDM)
```

# Post-estimation

## Predict and plot

```{r predict and plot, eval = FALSE}
Predictions <- predict(model, ...)
plot(Predictions)
```

# Model evaluation

## Spatial block

```{r spatBlock}
args(blockedCV)
```

## Leave-one-out

```{r LOO}
args(datasetOut)
```

# Quick example

## Introduction

-   We will primarily use the *Setophaga* dataset throughout this workshop

-   Contains information from three datasets:

    -   *eBird*

    -   *Pennsylvania Breeding Bird Atlas*

    -   *North American Breeding Bird Survey*

## `startISDM`

```{r startISDM2, eval = FALSE}
caerulescensModel <- startISDM(eBird, BBS, BBA, 
                       Boundary = PA,
                       Projection = proj, Mesh = mesh,
                       responsePA = 'NPres', 
                       responseCounts = 'Counts',
                       spatialCovariates = covariates,
              Formulas = list(
          covariateFormula = ~ elevation + I(elevation^2) + canopy + I(canopy^2))
                             )
```

## Specify Model

```{r specifyModel, eval = FALSE}
caerulescensModel$specifySpatial(sharedSpatial = TRUE,
                           prior.sigma = c(1, 0.1),
                           prior.range = c(15, 0.1))

caerulescensModel$addBias(datasetNames = 'eBird')
caerulescensModel$specifySpatial(Bias = TRUE, 
                           prior.sigma = c(1, 0.1),
                           prior.range = c(15, 0.1))

caerulescensModel$priorsFixed(Effect = 'Intercept',
                           mean.linear = 0, 
                           prec.linear = 0.1)
```

## `fitISDM`, `predict` and `plot`

```{r fitISDM2, eval = FALSE}
caerulescensEst <- fitISDM(data = caerulescensModel, 
                   options = modelOptions)

caerulescensPredictions <- predict(caerulescensEst,
                              data = fm_pixels(
                                     mesh = mesh,
                                     mask = PA),
                              spatial = TRUE,
                              n.samples = 1000)

plot(caerulescensPredictions, variable = c('mean', 'sd'))
```

## Next steps

-   Questions and discussions.

-   Pull *Github* repository: *PhilipMostert/PointedSDMs_Workshop.*

-   We will through *Vignette 2: Basic model* together.

## References

\tiny
