---
title: 'Presentation 3:'
subtitle: "Multi-Species Integrated Species Distribution Modeling in *PointedSDMs*\n\n'ISEC 2024 -- Swansea'"
author: "Ron R Togunov, Philip Mostert, Bob O'Hara"
institute: 'Norwegian University of Science and Technology'
date: '13 July 2024'
output: beamer_presentation
bibliography:  "`r here::here('References.bib')`"
biblio-style: authoryear
header-includes:
  - \let\verbatim\undefined
  - \let\verbatimend\undefined
  - \usepackage{listings}
  - \lstnewenvironment{verbatim}{\lstset{breaklines=true,basicstyle=\ttfamily\footnotesize}}{}
---

```{r Load PointedSDMs, echo = FALSE, warning = FALSE, message = FALSE}

library(knitr)
library(PointedSDMs)
opts_chunk$set(tidy.opts=list(width.cutoff=20),tidy=TRUE)

```

## Introduction to Multi-Species ISDMs

-   Extend single-species models to analyze multiple species simultaneously
-   Potential benefits:
    -   Improved estimates for rare species
    -   Better understanding of community-level patterns
    -   More efficient use of data

## The *Setophaga* Example

-   Three warbler species in Pennsylvania:
    -   *Setophaga fusca* (Blackburnian Warbler)
    -   *Setophaga caerulescens* (Black-throated Blue Warbler)
    -   *Setophaga magnolia* (Magnolia Warbler)

-   Data sources: eBird, BBS, BBA
-   Environmental covariates: Elevation, Canopy cover, Coniferous forest cover

## Model Initialization

```{r eval = FALSE}
speciesModel <- startSpecies(SetohagaData, 
                             Boundary = PA, 
                             Projection = proj, 
                             Mesh = mesh,
                             responsePA = 'NPres', 
                             trialsPA = 'Trials',
                             pointsSpatial = NULL,
                             spatialCovariates = covariates, 
                             speciesName = 'Species_name')
```

## Spatial Effect Structures

1. Independent (Replicate)
1. Shared
1. Copied


## Independent Spatial Effects
- Default multi-species model assumes each species has unique spatial field
  -  i.e., each treated as a `replicate`

```{r eval = FALSE}
startSpecies(..., speciesSpatial = "replicate", ...)
```

## Shared Spatial Effects
- If species are assumed to have a common spatial field, we can specify it is `shared`

```{r eval = FALSE}
startSpecies(..., speciesSpatial = "shared", ...)
```

## Copied Spatial Effects
```{r eval = FALSE}
startSpecies(..., speciesSpatial = "copy", ...)
```

## Species-Specific Covariate Effects
```{r eval = FALSE}
speciesModelConif <- startSpecies(..., 
                                  spatialCovariates = all_covariates, 
                                  speciesEnvironment = TRUE)
speciesModelConif$changeComponents(
  removeComponent = "Setophaga_caerulescens_conif_cover")
```

## Bias Correction in Multi-Species Models
```{r eval = FALSE}
speciesModel$addBias(datasetNames = 'eBird')
biasEst <- fitISDM(data = speciesModel, options = modelOptions)
plot(predict(biasEst, ..., bias = TRUE))
```

## Model Comparison
```{r eval = FALSE}
CopyspeciesEst$dic$dic
speciesSharedEst$dic$dic
speciesCopyEst$dic$dic
speciesEstConif$dic$dic
biasEst$dic$dic
```

## Ecological Implications
-  How do spatial effect structures affect species predictions?
-  What can we learn from species-specific covariate effects?
-  How does bias correction impact our understanding of species distributions?

## Next Steps

-  Questions and discussions
-  Hands-on exercise: Fitting multi-species models to your own data

## References

\tiny