---
title: 'Presentation 4:'
subtitle: "Multispecies Integrated Species Distribution Modeling in *PointedSDMs*\n\nISEC 2024 -- Swansea"
author: "Ron R. Togunov, Philip Mostert, Bob O'Hara"
institute: 'Norwegian University of Science and Technology'
date: '13 July 2024'
output: beamer_presentation
bibliography:  "`r here::here('References.bib')`"
biblio-style: authoryear
header-includes:
  - \let\verbatim\undefined
  - \let\verbatimend\undefined
  - \usepackage{listings}
  - \lstnewenvironment{verbatim}{\lstset{breaklines=true,basicstyle=\ttfamily\footnotesize}}{}
---

```{r Load PointedSDMs, echo = FALSE, warning = FALSE, message = FALSE}

library(knitr)
library(PointedSDMs)
opts_chunk$set(tidy.opts=list(width.cutoff=20),tidy=TRUE)

```

## Introduction to Multispecies ISDMs

::: incremental
-   Extend single-species models to analyze multiple species simultaneously
-   Potential benefits:
    -   Sharing information across species
    -   More efficient use of data
    -   Improved bias estimation
    -   Improved estimates (especially for rare species)
:::

## Multispecies model structure

::: incremental
-   Sharing spatial field & hyperparameters
    1.  Both shared across species 
       - Identical random field
    2.  Shared hyperparameters & independent field
       - Similar spatial structure, unique distributions
    3.  Neither are shared 
       - Unique spatial structure & distribution
-   Species-specific covariate effects
-   Dataset bias field
::: 


## Multispecies: Spatial field structure

- Initialise model
```{r eval = FALSE}
speciesModel <- startSpecies(SetohagaData, 
                             Boundary = PA, 
                             Projection = proj, 
                             Mesh = mesh,
                             responsePA = 'NPres', 
                             trialsPA = 'Trials',
                             spatialCovariates = covariates, 
                             speciesName = 'Species_name',
                             pointsSpatial = NULL,
                             speciesSpatial = ...)
```

1.  Shared both hyperparameters & field across species
    - `speciesSpatial = "share"`
2.  Shared hyperparameters & independent field (`intSDM` default)
    - `speciesSpatial = "replicate"`
3.  Independent hyperparameters & field  
    - `speciesSpatial = "copy"`

## Species-Specific Covariate Effects

-   Effect of covariate is specified uniquely across species
    - A covariate is applicable to some species
    - Test competing models hypothesis
    - Reduce overfitting 
    
```{r eval = FALSE}
# View components in current model
speciesModelConif$changeComponents()
# Remove specified components
speciesModelConif$changeComponents(removeComponent = "Setophaga_caerulescens_canopy")
```

## Bias Correction in Multispecies Models

-   Leverage information across species for improved bias estimation
```{r eval = FALSE}
speciesModel$addBias(datasetNames = 'eBird')
```

## Model Prediction
- Predicting only spatial effect
```{r eval = FALSE}
predict(..., spatial = TRUE)
```

- Predicting only covariate effect
```{r eval = FALSE}
predict(..., spatial = FALSE, covariates = model$spatCovs$name)
```

- Predicting covariate & spatial effects
```{r eval = FALSE}
predict(..., predictor = TRUE)
```

- Predicting bias field
```{r eval = FALSE}
predict(..., bias = TRUE)
```

## Supplementary activities

-   Compare various covariate prediction
-   Explore relationship between resolution/structure of covariates and INLA mesh

\tiny